\chapter*{Scope and Topic Map}\label{ch:auto-001}
\addcontentsline{toc}{chapter}{Scope and Topic Map}

\section*{How to Use This Text}\label{sec:auto-001}
Chapters 1 through 6 provide topic depth with definitions, equations, and worked examples.
Chapter 5 provides integrated review and cache-simulation context, while Chapter 6 introduces instruction-level parallelism and superscalar design ideas.

\section*{Topic Coverage Map}\label{sec:auto-002}
\paragraph{Chapter 1: Introduction}\label{par:auto-001}
Coverage includes speculation with and without prediction, grain sizes of parallelism (bit, instruction, thread, and inter-program), Flynn taxonomy, SIMD versus MIMD, pipelining as temporal parallelism, and the distinction between packed-word SIMD and pipelined vector operations.

\paragraph{Chapter 2: Performance Evaluation}\label{par:auto-002}
Coverage includes throughput versus runtime, the CPU-time equation in terms of IC, CPI, and cycle time, RISC versus CISC tradeoffs, Amdahl's and Gustafson's laws, and correct use of arithmetic versus harmonic means.

\paragraph{Chapter 3: Cache Memories}\label{par:auto-003}
Coverage includes SRAM versus DRAM and bit-line delay, locality, hierarchy levels, virtual-memory terminology and its analogy to caching, average access time, the three C's of misses, cache geometry and construction, replacement and write policies, techniques to improve miss rate, miss penalty, and hit time, and compiler/data-layout effects.

\paragraph{Chapter 4: Pipelining}\label{par:auto-004}
Coverage includes IF/ID/EX/MEM/WB behavior, speedup equations, CPI stall modeling, the principle that program behavior plus hazards produces stalls, structural and control hazards, delay and squash-slot ideas, BTB behavior with and without tags, return-address stacks, and direction prediction heuristics.

\paragraph{Chapter 5: Integrated Review and Cache Simulation Concepts}\label{par:auto-005}
Coverage includes homework-aligned synthesis, worked pipeline calculations, hierarchy semantics, prefetching concepts, and interpretation of simulator-style statistics and AAT formulas.

\paragraph{Chapter 6: ILP and Superscalar Architecture}\label{par:scope-ch6}
Coverage includes CPI-versus-IPC interpretation, historical superscalar context, fire/complete ordering models, scoreboarding and busy-bit scheduling, function-unit latency considerations, register renaming intuition, and speculation/retirement behavior in out-of-order-friendly pipelines.
