\chapter{Performance Evaluation}\label{ch:auto-003}


\section{Why Performance Evaluation Matters}\label{sec:auto-010}
Architecture decisions are only meaningful if they are evaluated with the right metric.
A design that improves one metric can degrade another, and a design that looks fast on one workload can be slow on the workload that actually matters.
The goal of this chapter is therefore methodological: define performance correctly, choose representative workloads, and use mathematically correct summaries when comparing alternatives.


\section{Throughput vs. Runtime}\label{sec:auto-011}
Two primary performance metrics appear repeatedly in computer architecture.
\textit{Runtime} (also called response time or turnaround time) measures how long one job takes from start to completion.
\textit{Throughput} measures how many jobs complete per unit time.
These metrics are related but not interchangeable.

For a single-user interactive application, runtime is usually the dominant concern: the user wants one operation to finish quickly.
For servers and shared systems, throughput is often dominant: the owner cares about how many requests the system can process per second.
The same machine can be better under one metric and worse under the other, so performance claims must always specify which metric is being optimized.

This distinction also explains why pipelining can be valuable even when individual operation latency does not fall dramatically.
A deeper pipeline can increase the number of completed operations per unit time, improving throughput, while leaving single-operation latency unchanged or slightly worse.


\section{CPU Runtime Equation}\label{sec:auto-012}
The fundamental central processing unit (CPU) time equation used throughout performance analysis is (Eq.~\eqref{eq:auto-002}):
\begin{equation}\label{eq:auto-002}
    \text{CPU time} = \text{Instruction Count (IC)} \times \text{Cycles per Instruction (CPI)} \times \text{Cycle Time (CT)}.
\end{equation}
Using clock frequency $f = 1/\text{CT}$, the same equation is (Eq.~\eqref{eq:auto-003}):
\begin{equation}\label{eq:auto-003}
    \text{CPU time} = \frac{\text{IC} \times \text{CPI}}{f}.
\end{equation}
This decomposition is useful because each factor is controlled by different mechanisms.
\textit{Instruction count} is influenced by algorithm choice, compiler quality, and instruction set architecture (ISA) expressiveness.
\textit{Cycles per instruction (CPI)} is influenced by microarchitecture, hazard behavior, memory hierarchy behavior, and available parallelism.
\textit{Cycle time} is influenced by circuit and pipeline timing constraints.

For instruction mixes, total cycles are computed directly (Eq.~\eqref{eq:auto-004}):
\begin{equation}\label{eq:auto-004}
    \text{Total cycles} = \sum_i \left(\text{IC}_i \times \text{CPI}_i\right), \qquad
    \text{CPI}_{avg} = \frac{\text{Total cycles}}{\text{Total instructions}}.
\end{equation}

\paragraph{Worked example}\label{par:auto-009}
Consider a workload that executes 76 loads at 6 cycles each, 30 stores at 4 cycles each, 300 adds at 1 cycle each, and 58 multiplies at 5 cycles each.
Then total cycles are
\begin{equation}\label{eq:auto-005}
    76\cdot6 + 30\cdot4 + 300\cdot1 + 58\cdot5 = 1166.
\end{equation}
At 2.2\,MHz, runtime is
\begin{equation}\label{eq:auto-006}
    \frac{1166}{2.2\times 10^6} \approx 5.30\times 10^{-4}\text{s} = 0.53\text{ms}.
\end{equation}
If multiply latency is reduced from 5 cycles to 3 cycles, new total cycles become
\begin{equation}\label{eq:auto-007}
    76\cdot6 + 30\cdot4 + 300\cdot1 + 58\cdot3 = 1050,
\end{equation}
which yields a speedup of
\begin{equation}\label{eq:auto-008}
    \frac{1166}{1050}\approx 1.11.
\end{equation}
The example illustrates a recurring theme: accelerating only one instruction class can help, but total improvement depends on how often that class occurs.

\paragraph{Worked machine-comparison example}\label{par:auto-009a}
Now consider three machines that all execute the same source-level program.
Machine J runs at 5.1 GHz with $CPI=3.0$.
Machine K runs at 5.5 GHz with $CPI=3.7$ and the same dynamic instruction count as J\@.
Machine G runs at 6.1 GHz with $CPI=3.2$ but requires $1.5\times$ as many dynamic instructions.
Using Eq.~\eqref{eq:auto-003}, the J-versus-K comparison is
\begin{equation}\label{eq:auto-008a}
    \frac{T_K}{T_J}=\frac{3.7/5.5}{3.0/5.1}\approx 1.146,
\end{equation}
so K is slower even though its clock is faster.
For J versus G, runtimes are proportional to
\begin{equation}\label{eq:auto-008c}
    T_J \propto \frac{1.0\times 3.0}{5.1}, \qquad
    T_G \propto \frac{1.5\times 3.2}{6.1}.
\end{equation}
Numerically,
\begin{equation}\label{eq:auto-008b}
    T_J \propto 0.588,\qquad
    T_G \propto 0.787.
\end{equation}
So the faster clock does not rescue Machine G\@: the larger dynamic instruction count dominates and Machine J is faster overall.
This is the central Chapter 2 discipline: compare complete CPU time, not any single factor in isolation.


\section{RISC vs. CISC Through the Performance Equation}\label{sec:auto-013}
\textit{Reduced instruction set computer (RISC)} and \textit{complex instruction set computer (CISC)} are best understood as design tradeoffs, not as a simple ``better/worse'' ranking.
Historically, CISC designs emphasized richer, denser instructions to reduce dynamic instruction count and code size.
RISC designs emphasized simpler instructions and more regular pipelines, often improving CPI and enabling higher clock rates.
In CPU-time terms, this is a direct trade (Eq.~\eqref{eq:auto-009}):
\begin{equation}\label{eq:auto-009}
    \text{IC}\downarrow \text{may come with} \text{CPI}\uparrow \text{or} \text{CT}\uparrow,\quad
    \text{or vice versa.}
\end{equation}

In modern systems the boundary is less rigid than textbook definitions suggest, and x86 implementations, for example, decode complex instructions into simpler internal micro-operations and then execute them with highly RISC-like back ends.
Arm designs, traditionally associated with RISC, also include increasingly rich instruction features to improve efficiency for modern workloads.
The practical conclusion is that ISA style should be evaluated by end-to-end performance, power, and software ecosystem impact rather than by labels alone.


\section{Amdahl's Law}\label{sec:auto-014}
\textit{Amdahl's Law} quantifies the speedup limit when only part of a workload is accelerated.
If fraction $f$ of execution time benefits from a speedup factor $s$, total speedup is
\begin{equation}\label{eq:auto-010}
    S_{Amdahl}=\frac{1}{(1-f)+\frac{f}{s}}.
\end{equation}
The limit as $s\to\infty$ is
\begin{equation}\label{eq:auto-011}
    S_{\text{upper}}=\frac{1}{1-f}.
\end{equation}
This bound is the key insight: the unaccelerated fraction eventually dominates.

\paragraph{Worked example}\label{par:auto-010}
Consider a case in which 80\% of runtime is parallelizable and the accelerated part can use 100{,}000 processors ideally ($s=100{,}000$). Then
\begin{equation}\label{eq:auto-012}
    S=\frac{1}{0.2+\frac{0.8}{100000}}\approx 5.0.
\end{equation}
Even enormous parallel resources cannot overcome a significant serial fraction.
This is why architectural proposals must report both the accelerated fraction and the achieved acceleration factor.

\paragraph{Maximum-speedup interpretation}\label{par:auto-010a}
When a problem states that a new mechanism speeds up ``25\% of a benchmark program,'' the 25\% must be interpreted as a fraction of the \emph{original execution time} of the whole benchmark unless stated otherwise.
If that accelerated fraction were made infinitely fast, the maximum possible speedup would be
\begin{equation}\label{eq:auto-012a}
    S_{\max}=\frac{1}{1-0.25}=\frac{4}{3}\approx 1.33.
\end{equation}
This is the limiting-case form of Amdahl's bound: once the improved portion becomes arbitrarily fast, the remaining 75\% of the original runtime dominates completely.


\section{Gustafson's Law and Why It Does Not Contradict Amdahl}\label{sec:auto-015}
\textit{Gustafson's Law} addresses a different question: instead of fixing problem size, it assumes the workload scales with available processors.
Let $\alpha$ be the sequential fraction measured on an $N$-processor run.
Scaled speedup is
\begin{equation}\label{eq:auto-013}
    S_{Gustafson}=\alpha + (1-\alpha)N.
\end{equation}
This often grows nearly linearly with $N$ when $\alpha$ is small.

Amdahl and Gustafson do not contradict each other because their assumptions differ.
Amdahl is \textit{strong-scaling} analysis (fixed total work).
Gustafson is \textit{weak-scaling} analysis (larger total work as resources grow).
Both are valid, but they answer different engineering questions.

\paragraph{Example}\label{par:auto-011}
With $\alpha=0.15$ and $N=16$,
\begin{equation}\label{eq:auto-014}
    S_{Gustafson}=0.15+0.85\cdot16=13.75.
\end{equation}
This can exceed Amdahl's fixed-size bound because the scaled workload includes more total useful work.


\section{Combining Performance Numbers Correctly}\label{sec:auto-016}
Single-number summaries are useful only when the averaging method matches the quantity being averaged.
For times (seconds per task), \textit{arithmetic mean} is appropriate (Eq.~\eqref{eq:auto-015}):
\begin{equation}\label{eq:auto-015}
    \bar{T}_{arith}=\frac{1}{n}\sum_{i=1}^{n}T_i.
\end{equation}
For rates (tasks per second), use \textit{harmonic mean} (Eq.~\eqref{eq:auto-016}):
\begin{equation}\label{eq:auto-016}
    \bar{R}_{harm}=\frac{n}{\sum_{i=1}^{n}\frac{1}{R_i}}.
\end{equation}

Using arithmetic mean on rates can select the wrong system.
Suppose machine A runs two workloads at 4 and 4 transactions/s, while machine B runs them at 2 and 7 transactions/s.
Arithmetic means are 4.0 and 4.5, which appears to favor B\@.
But for one transaction of each workload, total throughput is determined by total work divided by total time (Eq.~\eqref{eq:auto-017}):
\begin{equation}\label{eq:auto-017}
    R_A=\frac{2}{\frac{1}{4}+\frac{1}{4}}=4,\qquad
    R_B=\frac{2}{\frac{1}{2}+\frac{1}{7}}=\frac{28}{9}\approx 3.11.
\end{equation}
Under equal-work comparison, A is better.
This is exactly why rates should be combined harmonically, not arithmetically.

\paragraph{Weighted workload mixes}\label{par:auto-011a}
Real benchmark suites are often not uniform.
If workload $i$ contributes fraction $w_i$ of the jobs or requests, with $\sum_i w_i=1$, then the weighted average runtime is
\begin{equation}\label{eq:auto-016a}
    \bar{T}_{weighted}=\sum_i w_i T_i.
\end{equation}
For rates under the corresponding fixed-work interpretation, the comparable aggregate rate is
\begin{equation}\label{eq:auto-016b}
    R_{eq}=\frac{1}{\sum_i \frac{w_i}{R_i}}.
\end{equation}
These weighted forms are the correct way to combine benchmark results when workloads do not occur equally often.

\paragraph{Worked weighted-runtime example}\label{par:auto-011b}
Suppose three workloads occur with weights $0.50$, $0.30$, and $0.20$, and one machine runs them in 4 s, 9 s, and 6 s, respectively.
Then the weighted average runtime is
\begin{equation}\label{eq:auto-016c}
    \bar{T}_{weighted}=0.50(4)+0.30(9)+0.20(6)=5.9\ \text{s}.
\end{equation}
The same arithmetic form should \emph{not} be applied blindly to rates.
If the reported numbers were throughputs, the fixed-work interpretation would instead require the weighted harmonic form in Eq.~\eqref{eq:auto-016b}.


\section{Workload Dependence and Benchmarking}\label{sec:auto-017}
Performance numbers are workload-dependent.
Miss ratio, CPI, and even effective clock behavior under thermal constraints depend on the executed program mix.
A claim such as ``processor X is faster'' is incomplete unless it states the benchmark set, input sizes, and aggregation method.

Representative benchmarking therefore requires matching the measured workloads to expected deployment behavior.
A web service, a game engine, and a scientific simulation stress very different execution paths and memory behaviors.
Chapter 2 methodology demands that architects evaluate on relevant traces and report assumptions explicitly; otherwise conclusions will not transfer to real systems.


\section{Cost-Aware Performance Comparison}\label{sec:auto-017a}
Raw speedup is not always the only decision metric.
If machine $m$ costs $Cost_m$ and delivers speedup $S_m$ relative to a chosen baseline, two useful value measures are
\begin{equation}\label{eq:auto-017b}
    \text{speedup per dollar}=\frac{S_m}{Cost_m},
    \qquad
    \text{cost per unit speedup}=\frac{Cost_m}{S_m}.
\end{equation}
The first form is convenient when maximizing performance for a fixed budget.
The second form is convenient when minimizing cost for a required performance gain.
Both require the same discipline as every other Chapter 2 metric: use one baseline, one workload definition, and one aggregation method throughout the comparison.

\paragraph{Worked value-comparison example}\label{par:auto-011c}
Suppose Server A costs \$4{,}000 and delivers speedup 1.25, while Server B costs \$5{,}400 and delivers speedup 1.55.
Their speedup per dollar values are
\begin{equation}\label{eq:auto-017c}
    \frac{1.25}{4000}=3.125\times 10^{-4},\qquad
    \frac{1.55}{5400}\approx 2.87\times 10^{-4},
\end{equation}
so Server A gives slightly more performance per dollar.
Their cost per unit speedup values are
\begin{equation}\label{eq:auto-017d}
    \frac{4000}{1.25}=3200,\qquad
    \frac{5400}{1.55}\approx 3483.87,
\end{equation}
which leads to the same conclusion under the reciprocal metric.
This example is intentionally simple: the point is to show that cost-aware comparisons are still just ratio comparisons once a common baseline is fixed.


\section{Chapter 2 Summary}\label{sec:auto-018}
Chapter 2 provides the framework for disciplined performance reasoning.
Runtime and throughput are distinct objectives and must not be mixed casually.
CPU time decomposes into IC, CPI, and cycle time, which allows precise reasoning about optimization impact.
RISC and CISC are best evaluated as tradeoff points in Eq.~\eqref{eq:auto-002} rather than as absolute categories.
Amdahl's Law gives hard limits for fixed-size speedup, while Gustafson's Law explains scaled-workload growth without contradiction.
Finally, performance summaries require correct statistics: arithmetic means for time, harmonic means for rate, and weighted forms when the workload mix is not uniform.
These tools form the quantitative core used in later chapters.


\section{Practice Questions}\label{sec:ch2-practice}
The following questions reinforce the main ideas of Chapter 2 and are intended for self-checking after the summary.
\begin{enumerate}
    \item Machine A executes a workload with $IC=1.2\times 10^9$, $CPI=1.5$, and clock rate $2.4$ GHz. Machine B executes the same workload with $IC=0.9\times 10^9$, $CPI=1.9$, and clock rate $3.0$ GHz. Compute the runtime of both machines and the speedup of the faster machine over the slower one.
    \item Two machines execute the same program. Machine J runs at $5.1$ GHz with $CPI=3.0$. Machine K runs at $5.5$ GHz with $CPI=3.7$. Compute the speedup of K over J. Then consider Machine G, which runs at $6.1$ GHz with $CPI=3.2$ but requires $1.5\times$ as many dynamic instructions as J for the same work. Compute the speedup of G over J.
    \item A benchmark mix contains three workloads with weights $0.50$, $0.30$, and $0.20$. On one machine, their runtimes are 4 s, 9 s, and 6 s. Compute the weighted average runtime. Then explain why the same weighted arithmetic mean would be inappropriate if the reported quantities were throughputs instead of runtimes.
    \item A proposed optimization accelerates the branch-recovery portion of a program, which currently accounts for 30\% of total runtime, by a factor of 4. Use Amdahl's Law to compute the overall speedup. Then compute the maximum possible speedup if that portion were made infinitely fast.
    \item A program spends 20\% of its time in vectorizable code. Suppose an added accelerator makes only that portion run four times faster. Compute the overall speedup. Then explain why adding more parallel hardware gives diminishing returns when the accelerated fraction is fixed.
    \item A parallel machine runs a scaled problem with serial fraction $0.10$ on 16 processors. Compute the speedup predicted by Gustafson's Law. Then explain why this result does not contradict Amdahl's Law.
    \item For each of the following, state whether arithmetic mean, harmonic mean, or a weighted variant is the correct aggregation method, and explain why: runtimes for an equal-work benchmark suite, throughputs for an equal-work benchmark suite, and runtimes for a production workload mix with known fractions.
    \item Two candidate servers cost \$4{,}000 and \$5{,}400 and deliver speedups of 1.25 and 1.55 relative to the same baseline. Compute speedup per dollar and cost per unit speedup for both systems. State which system is preferable under each metric.
\end{enumerate}
