\chapter{Integrated Synthesis and Cache Simulation Concepts}\label{ch:auto-006}


\section{Chapter Purpose}\label{sec:auto-046}
This chapter draws together the main ideas from Chapters 1--4.
The goal is not to repeat those chapters in full, but to show how performance metrics, hazards, memory behavior, and simulation-oriented abstractions fit into a single analysis framework.


\section{Integrated Themes from Chapters 1 and 2}\label{sec:auto-047}
The first two chapters establish the measurement framework used throughout the rest of the book.
An architectural change should first be classified by what it is trying to improve: parallelism, speculation, or some combination of the two.
Its performance impact must then be evaluated with the correct metric, usually runtime for one job or throughput for a stream of jobs.
The central CPU-time equation,
\[
\text{CPU time}=\text{IC}\times\text{CPI}\times\text{CT},
\]
turns that reasoning into a concrete tradeoff among instruction count, pipeline efficiency, and clock period.
Within that framework, Amdahl's law bounds fixed-size speedup, Gustafson's law explains scaled-size behavior, and the choice between arithmetic and harmonic means determines whether a summary is mathematically meaningful.

\paragraph{Worked check}\label{par:auto-026}
Consider an optimization that accelerates 30\% of runtime by $2\times$. The total speedup is
\begin{equation}\label{eq:auto-045}
    S=\frac{1}{(1-0.3)+\frac{0.3}{2}}=\frac{1}{0.85}\approx 1.176.
\end{equation}
The unaccelerated fraction remains in the denominator and bounds overall speedup.

\paragraph{Worked check: maximum possible speedup}\label{par:auto-026a}
Consider a mechanism that accelerates 25\% of the original benchmark runtime and makes that portion arbitrarily fast. The maximum total speedup is
\begin{equation}\label{eq:auto-045a}
    S_{\max}=\frac{1}{1-0.25}=\frac{4}{3}\approx 1.33.
\end{equation}
This is the standard way to interpret the \emph{maximum} speedup of a partial optimization.


\section{Integrated Themes from Chapter 3}\label{sec:auto-048}
Chapter 3 adds the memory-hierarchy side of the performance model.
That means translating a cache specification such as $(C,B,S)$ into sets, ways, and tag/index/offset fields; distinguishing compulsory, conflict, and capacity misses; and understanding how replacement and write policies alter observed traffic.
It also means using
\[
AAT=HT+MR\cdot MP
\]
as a compact summary of hierarchy behavior while remembering that cache size, block size, and associativity can influence all three terms in different ways.

\paragraph{Worked check}\label{par:auto-027}
Consider a cache with $HT=1.8$ ns, $MR=0.06$, and $MP=28$ ns (Eq.~\eqref{eq:auto-046}):
\begin{equation}\label{eq:auto-046}
    AAT = 1.8 + 0.06\cdot 28 = 3.48\ \text{ns}.
\end{equation}

\paragraph{Worked check: repeated trace reasoning}\label{par:auto-027a}
Consider a direct-mapped L1 with $(C,B,S)=(8,4,0)$, write-back/write-allocate, and the repeated access sequence
\begin{equation}\label{eq:auto-046a}
    R\ 0x41F1,\ R\ 0xB54A,\ W\ 0x4204,\ W\ 0xCA3B,\ R\ 0x420E,\ W\ 0xB537.
\end{equation}
With $B=4$, the six accesses touch the five blocks $\{0x41F,0xB54,0x420,0xCA3,0xB53\}$ because $0x4204$ and $0x420E$ lie in the same block.
The first iteration therefore has 5 compulsory misses.
The key structural fact is that blocks $0xCA3$ and $0xB53$ map to the same direct-mapped set, so after warm-up the sequence continues to suffer two conflict misses per iteration.
Over iterations 2--10, that produces 18 additional misses.
Those 18 misses are the non-compulsory misses for the full run, and because the conflicting writes repeatedly evict dirty lines, the full sequence produces 19 writebacks if dirty lines are \emph{not} flushed after the final access.

\paragraph{Victim cache and +1 prefetch contrast}\label{par:auto-027b}
That same trace separates two common remedies cleanly.
A small victim cache can eliminate the recurring two-block ping-pong at one set; after warm-up, the combined L1+victim-cache miss count can drop to zero because each conflicting block is recovered from the victim structure instead of DRAM\@.
By contrast, a $+1$ prefetcher is not automatically a cure.
In this trace it can prefetch block $0x420$ early enough to remove one first-iteration demand miss, but it does not remove the recurring conflict between $0xCA3$ and $0xB53$.
This is the key design lesson of the trace: next-block prefetching and victim caching address different miss mechanisms.


\section{Integrated Themes from Chapter 4}\label{sec:auto-049}
Chapter 4 connects pipeline structure to observed performance.
Reading a timing diagram is therefore an exercise in inference: the reader is reconstructing hazards, bypass paths, memory behavior, and branch-handling policy from visible stage occupancy.
The same chapter also introduces the progression from static branch handling to dynamic predictors such as always taken, one-bit and Smith-counter schemes, GShare, GSelect, Yeh/Patt, perceptrons, and hybrid tournament predictors.

\subsection*{Pipeline Speedup}\label{subsec:auto-012}
Assume a 5-stage design where each stage has 250 ps logic delay and pipeline-register overhead is 30 ps.
Unpipelined instruction time is $5\times250=1250$ ps.
Pipelined clock is $250+30=280$ ps.
Ideal long-run speedup is (Eq.~\eqref{eq:auto-047}):
\begin{equation}\label{eq:auto-047}
    S\approx\frac{1250}{280}\approx 4.46.
\end{equation}
This is below 5 because register overhead is non-zero.

\subsection*{CPI with Branch Stalls}\label{subsec:auto-013}
If branch frequency is 18\% and each branch contributes an average of 1.6 stall cycles (Eq.~\eqref{eq:auto-048}):
\begin{equation}\label{eq:auto-048}
    CPI = 1 + 0.18\times1.6 = 1.288.
\end{equation}
This illustrates how even moderate branch penalties raise CPI above the ideal value of 1.

\subsection*{Software Branch Handling}\label{subsec:auto-013a}
When a branch has delay slots, every instruction moved into those slots must be safe regardless of whether the branch is taken.
When a branch instead has squash slots under an ``assume taken'' policy, the compiler may place instructions from the predicted taken path there, because wrong-path entries will be canceled.
The resulting placement rule is:
\begin{enumerate}
    \item fill delay slots only with path-independent instructions,
    \item fill squash slots from the predicted path when profitable, and
    \item leave \texttt{NOP}s when correctness would otherwise be uncertain.
\end{enumerate}


\section{Cache Simulation Conceptual Model}\label{sec:auto-050}
The cache simulation framework models hierarchy behavior at the level of addresses, tags, and policy events rather than payload data values.
A common parameterization is (Eq.~\eqref{eq:auto-049}):
\begin{equation}\label{eq:auto-049}
    (C_1,B,S_1,C_2,S_2,\{\text{MIP,LIP}\},\{\text{None,+1,Markov,Hybrid}\},r).
\end{equation}
In this notation, $C$ controls capacity ($2^C$ bytes), $B$ controls block size ($2^B$ bytes), and $S$ controls associativity ($2^S$ ways).
Here \textit{most recently used insertion policy (MIP)} and \textit{least recently used insertion policy (LIP)} are common insertion-policy options.


\section{Hierarchy Semantics in the Model}\label{sec:auto-051}
The hierarchy model treats L1 accesses separately from L2 internals, while L2 requests are viewed as arriving from an upper-level requester.
This separation makes it easier to reason about statistics at each level without conflating local policy with downstream timing.

Write strategy also changes the meaning of those statistics.
\textit{Write-back, write-allocate (WBWA)} behavior in L1 emphasizes dirty-state tracking and deferred propagation, whereas \textit{write-through, write-no-allocate (WTWNA)} behavior in L2 emphasizes immediate lower-level visibility and no allocation on write misses.

The ordering of read-miss repair and writeback events can also matter.
Different servicing orders change replacement-state evolution and therefore alter downstream hit, miss, and writeback counts.
If a lower level is disabled in the model, it is often best understood as an always-miss transit layer whose contribution to AAT is only the cost of passing requests to the next active level.


\section{Prefetching Concepts}\label{sec:auto-052}
\textit{$+1$ prefetching} predicts the next sequential block ($X+1$) after a demand miss on block $X$.
\textit{Markov prefetching} instead models miss-stream transitions as a finite-state table, associates each source miss block with a bounded set of likely successors and their frequencies, and selects a prediction from the highest-frequency successor under a deterministic tie-break rule.
\textit{Hybrid prefetching} combines correlation-based prediction with sequential fallback when correlation information is absent.

To interpret prefetch behavior correctly, the simulator must also distinguish demand-filled lines from prefetched lines.
A prefetch metadata bit provides that distinction.
The first demand use of a prefetched line contributes to prefetch-hit interpretation, while eviction before first demand use contributes to prefetch-miss interpretation.


\section{Statistics and AAT Interpretation}\label{sec:auto-053}
Typical hierarchy statistics include access counts, hit/miss counts, hit/miss ratios, writeback counts, and prefetch effectiveness counters.
In two-level analysis (Eq.~\eqref{eq:auto-050}):
\begin{equation}\label{eq:auto-050}
    AAT_{L1}=HT_{L1}+MR_{L1}\cdot MP_{L1},\qquad MP_{L1}=AAT_{L2}.
\end{equation}
\begin{equation}\label{eq:auto-051}
    AAT_{L2}=HT_{L2}+MR_{L2,read}\cdot DRAM\_TIME.
\end{equation}
\begin{equation}\label{eq:auto-052}
    DRAM\_TIME = DRAM\_AT + DRAM\_AT\_PER\_WORD\cdot WORDS\_PER\_BLOCK.
\end{equation}
This form emphasizes that higher-level penalty is the lower-level average access behavior, not a fixed constant.


\section{Conceptual Tradeoffs Across Parameters}\label{sec:auto-054}
Increasing $C$ generally reduces capacity misses but tends to raise hit time.
Increasing $S$ reduces conflict misses but increases lookup complexity.
Increasing $B$ can exploit spatial locality but may also increase pollution and miss penalty.
More aggressive prefetching can reduce compulsory misses yet increase bandwidth pressure and useless fills.


\section{Chapter 5 Summary}\label{sec:auto-055}
Chapter 5 draws together the earlier performance, pipeline, and memory ideas into one simulation-oriented view.
Pipeline timing and cache behavior are not separate topics; they are two views of the same broader question of where cycles are spent and why.
The simulation concepts in this chapter provide a concrete way to observe those tradeoffs in controlled workloads.
That same simulator-oriented reasoning extends naturally to branch-prediction accuracy, fetched-versus-retired instruction counts, queue pressure, and reorder-buffer pressure, which are developed further in Chapter 6.
A grouped formula index for the full text is provided in Appendix~\ref{ch:formula-compendium}.
