\chapter[Dynamic ILP and Out-of-Order]{Dynamic ILP: Superscalar Scheduling and Out-of-Order Execution}\label{ch:ilp-superscalar}

\section{Why Cycles per Instruction Can Go Below One}\label{sec:ilp-cpi-ipc}
In basic single-issue pipelines, designers often treat cycles per instruction (CPI) near 1 as a practical lower bound.
Superscalar execution challenges that assumption by allowing multiple instructions to make progress in the same cycle.
This broader goal is usually described as instruction-level parallelism (ILP).
In this context, instructions per cycle (IPC) is the natural companion metric, as shown in Eq.~\eqref{eq:ilp-ipc-cpi}:
\begin{equation}\label{eq:ilp-ipc-cpi}
    IPC=\frac{1}{CPI}.
\end{equation}
So achieving $CPI<1$ is exactly equivalent to achieving $IPC>1$.
A useful way to view design limits is that average retire throughput cannot exceed the average throughput of the stages before it (Eq.~\eqref{eq:ilp-throughput-bound}):
\begin{equation}\label{eq:ilp-throughput-bound}
    IPC_{retire} \le \min\!\left(IPC_{fetch},\ IPC_{dispatch},\ IPC_{execute},\ IPC_{state\ update}\right).
\end{equation}
If any major stage can only handle one instruction per cycle on average, then the whole machine cannot retire two instructions per cycle on average, even if another stage is wide enough to do so.
In other words, the minimum term in Eq.~\eqref{eq:ilp-throughput-bound} is the throughput bottleneck.

Each term corresponds to a different possible limit.
$IPC_{fetch}$ asks how many useful instructions the front end can supply per cycle.
$IPC_{dispatch}$ asks how many decoded instructions can be prepared and sent onward.
$IPC_{execute}$ asks how many instructions the function units (FUs) and memory system can actually complete.
$IPC_{state\ update}$ asks how quickly the machine can accept completed work into precise architectural state.
The retire rate cannot exceed the weakest of these rates, because retirement is the final visible outcome of all the earlier pipeline activity.


\section{Historical Motivation}\label{sec:ilp-history}
Historically, supercomputer scalar units showed that out-of-order mechanisms could deliver $IPC>1$ long before that became mainstream in commodity processors.
The central lesson is that high throughput comes from combining several mechanisms rather than relying on any one of them in isolation.
Stalls are reduced through coordinated fetch, scheduling, execution, and retirement policies.
Modern superscalar designs extend the same idea by combining hazard management with wider issue and stronger speculation.


\section{Superscalar Pipeline View}\label{sec:ilp-pipeline-model}
A practical superscalar model can be described with five main logical units plus a final retirement or commit action.
Figure~\ref{fig:ilp-superscalar-pipeline} provides a compact visual summary of how these units interact in the overall datapath.
In that figure, the boxed labels denote the main structures or stages, while the smaller labels on the right identify the dominant action on the adjacent arrow bundle: fetch, dispatch, fire or issue, complete, and retire or commit.
The front end is drawn as separate Branch Handling and Instruction Fetch boxes, but together they still represent the single logical instruction fetch (IF) unit described below.

\paragraph{1. Branch Handling and Instruction Fetch (IF)}\label{par:ilp-fetch}
This unit performs instruction fetch (IF) while consulting branch-handling structures such as direction predictors, branch-target mechanisms, and return-address prediction.
Its goal is to produce multiple candidate instructions per cycle.

\paragraph{2. Dispatch and Decode (DISP)}\label{par:ilp-dispatch}
This unit decodes incoming instructions in program order and dispatches them toward the scheduling structures.
Dispatch is where program order is still explicit, which matters for precise recovery.

\paragraph{3. Scheduling}\label{par:ilp-schedule}
Scheduling assigns ready instructions to available FUs in time.
This is where execution can diverge from program order to exploit parallelism.

\paragraph{4. Execute (EX)}\label{par:ilp-execute}
FUs perform arithmetic, memory, and control operations with operation-specific latencies.
Multiple FUs allow parallel progress.

\paragraph{5. State Update (SU)}\label{par:ilp-state-update}
State update gathers completed results, handles squash or recovery, and prepares instructions for precise architectural completion.
This unit supports squash/recovery when speculation is wrong.
The final outgoing flow from this unit is the retirement or commit action, where results become architecturally visible in program order.
In Fig.~\ref{fig:ilp-superscalar-pipeline}, retirement is represented by the final outgoing arrows below State Update rather than by a separate boxed stage.

\begin{figure}[t]
    \centering
    \resizebox{\textwidth}{!}{%
    \begin{tikzpicture}[
        font=\small,
        >=Latex,
        box/.style={draw, rectangle, minimum height=12mm, minimum width=34mm, align=center},
        smallbox/.style={draw, rectangle, minimum height=9.5mm, minimum width=16mm, align=center},
        queue/.style={draw, rectangle, minimum height=7.5mm, minimum width=28mm, draw=red!65!black},
        stagearrow/.style={->, thick},
        sidearrow/.style={->, semithick},
        dasharrow/.style={->, dashed, semithick}
    ]

    % Top row
    \node[box, minimum width=24mm] (branch) at (0,0) {Branch\\Handling};
    \node[box, minimum width=52mm] (fetch) at (4.6,0) {Instruction Fetch};
    \node[smallbox] (icache) at (8.6,0) {I\$};

    % Main pipeline
    \node[box, minimum width=62mm] (dispatch) at (4.6,-2.6) {Dispatch and Decode};
    \node[box, minimum width=62mm] (schedule) at (4.6,-5.0) {Schedule};
    \node[box, minimum width=68mm, minimum height=24mm] (execute) at (4.6,-8.1) {};
    \node[box, minimum width=68mm] (state) at (4.6,-11.4) {State Update};
    \node at ($(execute.north)+(0,-0.55)$) {Execute unit};

    % Functional units inside execute
    \node[smallbox, minimum width=10mm, minimum height=9.5mm] (fu1) at (3.0,-8.75) {FU};
    \node[smallbox, minimum width=10mm, minimum height=9.5mm] (fu2) at (4.2,-8.75) {FU};
    \node at (5.25,-8.75) {$\cdots$};
    \node[smallbox, minimum width=10mm, minimum height=9.5mm] (fu3) at (6.2,-8.75) {FU};

    % Side structures
    \node[box, minimum width=22mm, minimum height=24mm] (regfile) at (-1.3,-3.8) {Reg\\File};
    \node[smallbox, minimum width=14mm, minimum height=11mm] (dcache) at (9.2,-8.1) {D\$};

    % Queues on right
    \node[queue] (dqueue) at (10.9,-2.6) {};
    \node[queue] (squeue) at (10.9,-5.0) {};

    % Connections at top
    \draw[<->, thick] (branch.east) -- (fetch.west);
    \draw[<->, thick] (fetch.east) -- (icache.west);

    % Vertical flow
    \draw[stagearrow] ($(fetch.south)+(-1.6,0)$) -- ($(dispatch.north)+(-1.6,0)$);
    \draw[stagearrow] ($(fetch.south)+(-0.4,0)$) -- ($(dispatch.north)+(-0.4,0)$);
    \draw[stagearrow] ($(fetch.south)+(1.6,0)$) -- ($(dispatch.north)+(1.6,0)$);
    \node at (5.2,-1.3) {$\cdots$};

    \draw[stagearrow] ($(dispatch.south)+(-1.6,0)$) -- ($(schedule.north)+(-1.6,0)$);
    \draw[stagearrow] ($(dispatch.south)+(-0.4,0)$) -- ($(schedule.north)+(-0.4,0)$);
    \draw[stagearrow] ($(dispatch.south)+(1.6,0)$) -- ($(schedule.north)+(1.6,0)$);
    \node at (5.2,-3.8) {$\cdots$};

    \draw[stagearrow] ($(schedule.south)+(-1.6,0)$) -- ($(execute.north)+(-1.6,0)$);
    \draw[stagearrow] ($(schedule.south)+(-0.4,0)$) -- ($(execute.north)+(-0.4,0)$);
    \draw[stagearrow] ($(schedule.south)+(1.6,0)$) -- ($(execute.north)+(1.6,0)$);
    \node at (5.2,-6.4) {$\cdots$};

    \draw[stagearrow] (fu1.south) -- ($(state.north)+(-1.6,0)$);
    \draw[stagearrow] (fu2.south) -- ($(state.north)+(-0.4,0)$);
    \draw[stagearrow] (fu3.south) -- ($(state.north)+(1.6,0)$);

    \draw[stagearrow] ($(state.south)+(-1.6,0)$) -- ++(0,-0.8);
    \draw[stagearrow] ($(state.south)+(-0.4,0)$) -- ++(0,-0.8);
    \draw[stagearrow] ($(state.south)+(1.6,0)$) -- ++(0,-0.8);
    \node at (5.2,-12.6) {$\cdots$};

    % Left-side register file connections
    \draw[sidearrow] (regfile.north east) -- ($(dispatch.west)+(0,0.00)$);
    \draw[sidearrow] (regfile.south east) -- ($(schedule.west)+(0,0.00)$);
    \draw[dasharrow] ($(state.west)+(-0.2,0)$) -- ++(-2.3,0) -- ++(0,5.0) -- (regfile.south);

    % D-cache connection
    \draw[-, semithick] (execute.east) -- (dcache.west);

    % Labels on right
    \node[anchor=west] at (7.0,-1.3) {fetch};
    \node[anchor=west, text=red!65!black] at (12.4,-2.6) {Dispatch Queue};
    \node[anchor=west] at (7.0,-3.8) {dispatch};
    \node[anchor=west, text=red!65!black] at (12.4,-5.0) {Scheduling Queue};
    \node[anchor=west] at (7.0,-6.4) {fire or issue};
    \node[anchor=west] at (7.0,-10.0) {complete};
    \node[anchor=west] at (7.0,-12.3) {retire or commit};

    \end{tikzpicture}
    }
    \caption{Logical superscalar pipeline view illustrating one representative organization. The figure is not specific to FICI, FICO, or FOCO; those classifications depend on the firing and completion policy. Boxed elements denote major structures or stages, while the right-side annotations denote the adjacent flow actions.}
    \label{fig:ilp-superscalar-pipeline}
\end{figure}


\section{Dependency Types and Why They Matter}\label{sec:ilp-dependencies}
Superscalar scheduling must respect three dependency classes.
\begin{itemize}
    \item \textit{read-after-write (RAW)} is a true flow dependency.
    A later instruction needs a value produced by an earlier one, so executing them in the wrong order would use stale data and produce the wrong result.
    \item \textit{write-after-read (WAR)} is an anti-dependency.
    A later instruction must not overwrite a register before an earlier instruction has finished reading it, even though the two instructions do not share a true data value.
    \item \textit{write-after-write (WAW)} is an output dependency.
    Two instructions write the same destination, so their writes must appear in the correct order or the final architectural value will be wrong.
\end{itemize}
Only RAW is fundamentally unavoidable.
WAR and WAW are name dependencies that arise from register reuse and can be reduced or removed by renaming the register.


\section{Fire/Complete Taxonomy}\label{sec:ilp-fire-complete}
A compact way to classify designs is by fire order and complete order.
\begin{itemize}
    \item \textit{fire in order, complete in order (FICI)} is the classic simple-pipeline behavior.
    Instructions begin execution in program order and also finish in program order, which simplifies control but limits how much parallelism the machine can exploit.
    \item \textit{fire in order, complete out of order (FICO)} is a simpler superscalar step.
    Instructions still begin execution in order, but different latencies allow later instructions to finish first, so some parallel speedup is possible without fully out-of-order scheduling.
    \item \textit{fire out of order, complete out of order (FOCO)} is more aggressive and offers higher potential ILP\@.
    The scheduler may start and finish instructions whenever operands and hardware are ready, which exposes more ILP but requires more machinery to preserve correct architectural behavior.
\end{itemize}


\section{Scoreboarding Intuition (FICO Style)}\label{sec:ilp-scoreboard}
A scoreboard tracks whether source operands, destinations, and required function units are currently available.
A simplified FICO-style fire condition is shown in Eq.~\eqref{eq:ilp-fire-condition}.
\begin{equation}\label{eq:ilp-fire-condition}
    \text{Fire if }\neg busy(src_1)\ \wedge\ \neg busy(src_2)\ \wedge\ \neg busy(dest)\ \wedge\ \neg busy(FU_{needed}).
\end{equation}
When an instruction fires, destination and function-unit busy state are set.
When the instruction completes, those busy states are cleared according to the implementation timing rules.
The result is correct execution with moderate hardware complexity, but some parallel opportunities remain unused compared with more aggressive out-of-order firing.


\section{Function Units and Latency}\label{sec:ilp-latency}
Superscalar benefit depends on the latency mix and on the number of available units.
In many designs, arithmetic logic unit (ALU) operations have short latency, multiply operations take more cycles, and memory accesses vary widely depending on cache behavior.
The key point is that latency and throughput are related but not identical.
An operation may have a long latency and still support good throughput if the corresponding FU is pipelined and can accept a new instruction every cycle.
By contrast, a long-latency operation on a scarce or non-pipelined FU can create backpressure quickly, because younger ready instructions may have no place to execute.
Memory operations are especially important here.
A cache hit may behave like a modest fixed-latency operation, while a cache miss can hold resources for many extra cycles and expose whether the machine can continue doing useful independent work during the delay.

Unit count matters for the same reason.
If the machine has only one ALU, one multiplier, and one load/store path, the scheduler has few choices.
Adding more execution resources, including additional FUs such as ALUs, multipliers, or more load/store capacity, gives the scheduler more flexibility to overlap different instruction classes and reduce structural conflicts.
Designers therefore optimize both unit availability and scheduling policy, not just raw clock frequency.
The practical goal is not merely to make individual operations fast, but to ensure that the mix of latencies, FU counts, and pipeline structure supports high average IPC on realistic workloads.
A common simplified latency model is shown in Table~\ref{tab:ilp-typical-latencies}.
\begin{table}[t]
    \centering
    \begin{tabular}{|l|c|p{2.7in}|}
        \hline
        Type & Typical latency & Why \\
        \hline
        Floating-point (FP) multiply & 3 & Denormalize, multiply, renormalize. \\
        \hline
        FP add & 2 & Denormalize and add, then renormalize. \\
        \hline
        Load & 2 & Typical cache pipeline behavior. \\
        \hline
        Store & 1 & Write buffering, or a load/store buffer, can hide much of the visible latency. \\
        \hline
        Divide & Variable & Often implemented with unpipelined iterative methods such as Newton iteration, and many simplified models treat divide as too rare to emphasize. \\
        \hline
        ALU & 1 & Most common operation, so short latency is especially valuable. \\
        \hline
    \end{tabular}
    \caption{Typical simplified function-unit latencies used in superscalar examples.}
    \label{tab:ilp-typical-latencies}
\end{table}
In this simplified performance model, the exact cycle count for divide is not emphasized because divide is used relatively rarely.
In real designs, however, divide often has high latency and is less deeply pipelined than common ALU operations.


\section{Worked FICO Timing Example}\label{sec:ilp-fico-example}
The following seven-instruction sequence illustrates the scoreboard idea in a concrete schedule:
\begin{align*}
    A&:\ \texttt{ADD}\ R1,R2,R3 \\
    B&:\ \texttt{MUL}\ R5,R1,R4 \\
    C&:\ \texttt{ADD}\ R3,R1,R2 \\
    D&:\ \texttt{ADD}\ R1,R5,R4 \\
    E&:\ \texttt{MUL}\ R6,R7,R8 \\
    F&:\ \texttt{ADD}\ R1,R3,R3 \\
    G&:\ \texttt{ADD}\ R4,R3,R5
\end{align*}
Assume a fetch rate of two instructions per cycle, one ALU with latency one cycle, one multiplier with latency three cycles, and a simplified \textit{state update (SU)} stage used to make completion visible.
Table~\ref{tab:ilp-fico-timing} shows one compact timing schedule for this example.
The stage labels are \textit{instruction fetch (IF)}, \textit{dispatch (DISP)}, \textit{execute (EX)}, and \textit{state update (SU)}.
Repeated DISP entries indicate that the instruction is blocked in the in-order fire logic, not that it is being redispatched from scratch.
\begin{table}[t]
    \centering
    \setlength{\tabcolsep}{4pt}
    \renewcommand{\arraystretch}{1.15}
    \resizebox{\textwidth}{!}{%
    \begin{tabular}{c l | c c c c c c c c c c}
        \hline
        Instr.
        & Operation & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 \\
        \hline
        A & ADD $R1, R2, R3$ & IF & DISP & EX & SU &  &  &  &  &  &  \\
        B & MUL $R5, R1, R4$ & IF & DISP & DISP & EX & EX & EX & SU &  &  &  \\
        C & ADD $R3, R1, R2$ &  & IF & DISP & EX & SU &  &  &  &  &  \\
        D & ADD $R1, R5, R4$ &  & IF & DISP & DISP & DISP & DISP & EX & SU &  &  \\
        E & MUL $R6, R7, R8$ &  &  & IF & DISP & DISP & DISP & EX & EX & EX & SU \\
        F & ADD $R1, R3, R3$ &  &  & IF & DISP & DISP & DISP & DISP & EX & SU &  \\
        G & ADD $R4, R3, R5$ &  &  &  & IF & DISP & DISP & DISP & DISP & EX & SU \\
        \hline
    \end{tabular}}
    \caption{Timing table for a simplified FICO schedule. Repeated DISP entries indicate that an instruction remains blocked in the in-order dispatch-and-fire path while waiting for data or hardware to become available.}
    \label{tab:ilp-fico-timing}
\end{table}

\paragraph{Walkthrough}\label{par:ilp-fico-example-walkthrough}
Cycles 0--3 establish the basic pattern.
A and B are fetched together, but B cannot leave DISP immediately because it needs the value of $R1$ from A\@.
Once A executes, B can move forward, and C can also reach EX shortly afterward because it depends on the same produced value.
This is why B spends an extra cycle in DISP in Table~\ref{tab:ilp-fico-timing}: the machine is still firing instructions in program order, so the scoreboard must wait until A has released the needed state.

Cycles 4--6 expose the main weakness of FICO\@.
D is genuinely blocked because it needs $R5$ from B, so its repeated DISP entries reflect a real RAW dependence.
E, however, is independent of D and could use the multiplier, yet it still cannot fire early because the machine will not let younger instructions pass an older blocked instruction in the dispatch path.
That is the key limitation of fire-in-order scheduling: some stalls come from true dependencies, but others come from the policy itself.

Cycles 7--9 show both overlap and continued restrictions.
E eventually occupies the multiplier while F and G reach the ALU later, so the machine does achieve some parallel execution and some out-of-order completion.
At the same time, the single ALU and single multiplier keep the scheduler's choices narrow, and the in-order fire rule prevents the machine from exploiting all available independence.
The example therefore illustrates why FICO is attractive as a first superscalar step---it is much simpler than full FOCO---but also why more aggressive out-of-order firing and renaming are needed to capture more ILP\@.


\section{Register Renaming as the Next Step}\label{sec:ilp-renaming}
\textit{Register renaming} is the introduction of additional physical register identities so later writes do not falsely block earlier reads or writes.
Conceptually, renaming transforms many WAR/WAW hazards into non-hazards while preserving program meaning.
The dependency pressure that remains is primarily RAW, which reflects true data flow.
In a typical implementation, the registers named by the instruction set architecture (ISA) are the \textit{architected registers}, while the larger hidden pool used internally by the machine is the \textit{physical register file}.
The architected registers are the names the programmer sees in assembly language and that the ISA defines as part of the visible machine state.
The physical register file contains additional hidden storage used internally by the processor to hold renamed versions of those values.

This distinction matters because several in-flight instructions may all refer to the same architected destination at different moments in time.
Rather than repeatedly reusing one architectural name and creating false WAR/WAW conflicts, the machine can assign a different physical register to each new write while still preserving the illusion of a small fixed architected register set.
Older instructions can continue reading the older physical version, while younger instructions are redirected to the newer one.
A \textit{register alias table (RAT)} maps each architected register name to the physical register that currently holds its most recent value.
When dispatch renames a destination, it allocates a fresh physical register and updates the RAT so younger instructions will refer to the new location rather than the older one.
Ready bits in the physical register file then indicate whether each renamed value has actually been produced yet.
Some architectures describe the same underlying status with \textit{busy bits} instead.
The terminology varies, but the idea is the same: one convention marks a value as ready to use, while the other marks it as still busy or unavailable.
% Figure~\ref{fig:ilp-register-renaming} shows one concrete sequence for architected register \texttt{R1}.
% Instruction \texttt{I1} allocates a new physical destination \texttt{P8}, instruction \texttt{I2} captures that same \texttt{P8} as its source, and a later writer \texttt{I3} allocates yet another destination \texttt{P9}.
% Because \texttt{I2} already refers to \texttt{P8}, the later write to \texttt{P9} does not interfere with the older read, so the false write-after-read (WAR) and write-after-write (WAW) hazards disappear while the true read-after-write (RAW) dependence from \texttt{I1} to \texttt{I2} remains.
%
% \begin{figure}[t]
%     \centering
%     \resizebox{\textwidth}{!}{%
%     \begin{tikzpicture}[
%         font=\small,
%         >=Latex,
%         panel/.style={draw, rounded corners, fill=gray!4, minimum width=3.6cm, minimum height=6.6cm},
%         titlebox/.style={draw, rounded corners, fill=black!5, minimum width=3.0cm, minimum height=0.8cm, align=center, font=\bfseries},
%         instr/.style={draw, rounded corners, fill=blue!8, minimum width=3.0cm, minimum height=0.95cm, align=center},
%         ratbox/.style={draw, rounded corners, fill=yellow!15, minimum width=2.8cm, minimum height=0.9cm, align=left},
%         pregready/.style={draw, rounded corners, fill=green!12, minimum width=2.6cm, minimum height=0.75cm, align=left},
%         pregwait/.style={draw, rounded corners, fill=orange!18, minimum width=2.6cm, minimum height=0.75cm, align=left},
%         note/.style={draw, rounded corners, fill=white, minimum width=3.0cm, minimum height=1.0cm, align=left},
%         flow/.style={->, thick},
%         dep/.style={->, thick, green!50!black},
%         panelarrow/.style={->, semithick}
%     ]
%     \node[panel] (p0) at (0,0) {};
%     \node[panel] (p1) at (4.6,0) {};
%     \node[panel] (p2) at (9.2,0) {};
%     \node[panel] (p3) at (13.8,0) {};
%
%     \node[titlebox] at (0,2.55) {Step 0\\Before Dispatch};
%     \node[titlebox] at (4.6,2.55) {Step 1\\Dispatch \texttt{I1}};
%     \node[titlebox] at (9.2,2.55) {Step 2\\Dispatch \texttt{I2}};
%     \node[titlebox] at (13.8,2.55) {Step 3\\Dispatch \texttt{I3}};
%
%     \node[instr] at (0,1.25) {\texttt{R1} currently means\\the committed value};
%     \node[ratbox] (rat0) at (0,-0.1) {\textbf{RAT}\\\texttt{R1 -> P1}};
%     \node[pregready] (p0a) at (0,-1.45) {\textbf{P1}: old \texttt{R1}\\ready = 1};
%     \node[note] at (0,-2.85) {Only one visible name\\exists in the architected state.};
%
%     \node[instr] (i1) at (4.6,1.25) {\texttt{I1: ADD R1, R2, R3}\\allocate new destination};
%     \node[ratbox] (rat1) at (4.6,-0.1) {\textbf{RAT}\\\texttt{R1 -> P8}};
%     \node[pregready] at (4.6,-1.25) {\textbf{P1}: old \texttt{R1}\\ready = 1};
%     \node[pregwait] (p8a) at (4.6,-2.15) {\textbf{P8}: dest of \texttt{I1}\\ready = 0};
%     \node[note] at (4.6,-3.35) {A fresh physical register\\holds the next version of \texttt{R1}.};
%     \draw[flow] (i1.south) -- (p8a.north);
%
%     \node[instr] (i2) at (9.2,1.25) {\texttt{I2: SUB R4, R1, R5}\\capture source for \texttt{R1}};
%     \node[ratbox] (rat2) at (9.2,-0.1) {\textbf{RAT}\\\texttt{R1 -> P8}};
%     \node[pregwait] (p8b) at (9.2,-1.35) {\textbf{P8}: produced by \texttt{I1}\\ready = 0};
%     \node[note] (n2) at (9.2,-2.95) {\texttt{I2} records source\\\texttt{R1} as \texttt{P8}.\\This keeps the true RAW edge.};
%     \draw[flow] (rat2.south) -- (p8b.north);
%     \draw[dep] (p8b.south) -- (n2.north);
%
%     \node[instr] (i3) at (13.8,1.25) {\texttt{I3: MUL R1, R6, R7}\\allocate another destination};
%     \node[ratbox] (rat3) at (13.8,-0.1) {\textbf{RAT}\\\texttt{R1 -> P9}};
%     \node[pregready] (p8c) at (13.8,-1.2) {\textbf{P8}: older \texttt{R1} version\\used by \texttt{I2}};
%     \node[pregwait] (p9c) at (13.8,-2.15) {\textbf{P9}: dest of \texttt{I3}\\ready = 0};
%     \node[note] at (13.8,-3.35) {\texttt{I2} still reads \texttt{P8},\\while younger users of \texttt{R1}\\now see \texttt{P9}.};
%     \draw[flow] (i3.south) -- (p9c.north);
%     \draw[flow] (rat3.south) -- (p9c.north west);
%
%     \draw[panelarrow] (1.95,0) -- (2.65,0);
%     \draw[panelarrow] (6.55,0) -- (7.25,0);
%     \draw[panelarrow] (11.15,0) -- (11.85,0);
%
%     \node[draw, rounded corners, fill=green!8, minimum width=15.2cm, minimum height=1.2cm, align=left] (footer) at (6.9,-5.0) {
%         \textbf{Effect:} write-after-read (WAR) and write-after-write (WAW) hazards disappear because the later write gets a new physical name.
%         Only the real read-after-write (RAW) dependence remains: when \texttt{I1} finishes, it marks \texttt{P8} ready, and \texttt{I2} can wake up and use that value.
%     };
%     \end{tikzpicture}}
%     \caption{Register renaming over time.
%     The register alias table (RAT) always points to the newest physical version of an architected register, while older in-flight instructions keep the physical register numbers they captured earlier.
%     This removes false name dependencies while preserving true data flow.}
%     \label{fig:ilp-register-renaming}
% \end{figure}


\section{Tomasulo's Algorithm}\label{sec:ilp-tomasulo}
\textit{Tomasulo's algorithm} is a classic out-of-order execution method that implements \textit{FOCO} by allowing instructions to fire when their operands are ready rather than strictly in program order.
Its central idea is to combine renaming with \textit{reservation stations}, which are scheduling entries that hold an instruction until the required operands and execution resources are available.
Historically, this approach is associated with Bob Tomasulo's work on the IBM System/360 Model 91, and it remains the conceptual foundation behind many modern out-of-order schedulers.

\paragraph{Main structures}\label{par:ilp-tomasulo-structures}
A simplified Tomasulo-style machine contains a dispatch unit, a set of reservation stations, one or more FUs, a result-broadcast network, and a renamed register file.
Dispatch still runs in program order, but once instructions occupy reservation stations they may wake up and fire out of order.
Each reservation-station entry records the required FU type, the destination tag or physical register, and for each source either a ready value or a tag identifying which older instruction will eventually produce that value.

\paragraph{Dispatch, fire, and complete}\label{par:ilp-tomasulo-flow}
At dispatch time, each source operand is examined.
If the needed value is already ready, the value is copied into the reservation station.
If it is not ready, the reservation station records the producing tag instead.
The destination is renamed so younger instructions refer to the new producer rather than to the older architectural name.
Later, when both sources are ready and a matching FU is available, the instruction can fire even if an older unrelated instruction is still waiting somewhere else in the machine.
Completion is a separate event from firing: when an instruction completes, it broadcasts its tag and value on the common data bus (CDB) so dependent instructions can wake up and the renamed register file can be updated.

\paragraph{Illustrative four-instruction trace}\label{par:ilp-tomasulo-trace}
The following short trace captures the key behavior.
Instruction \textit{A} is a load that misses in the cache.
Instruction \textit{B} is an add that depends on \textit{A}\@.
Instruction \textit{C} is an independent load that hits in the cache.
Instruction \textit{D} is an add that depends on \textit{C}\@.
Because the machine is out of order and the cache is non-blocking, \textit{C} and then \textit{D} can make progress while \textit{A} is still waiting on the miss.
Figure~\ref{fig:ilp-tomasulo-cycle-table} gives a compact cycle-by-cycle reference for the same sequence before the text walks through it in words.
Each cell shows the dominant structure label first---dispatch (DISP), reservation station (RS), execute (EX), or common data bus (CDB)---followed by the local action.
\begin{figure}[t]
    \centering
    \setlength{\tabcolsep}{4pt}
    \renewcommand{\arraystretch}{1.2}
    \newcommand{\stagecell}[2]{\shortstack[c]{\rule{0pt}{2.9ex}\textbf{#1}\\[-0.1ex]\strut #2}}
    \newcommand{\rowlabel}[1]{\shortstack[c]{\rule{0pt}{2.5ex}\textbf{#1}\\[-0.1ex]\rule[-0.9ex]{0pt}{2.0ex}}}
    \resizebox{\textwidth}{!}{%
    \begin{tabular}{|>{\raggedright\arraybackslash}m{28mm}|c|c|c|c|c|c|c|}
        \hline
        \textbf{Instruction / event} & \textbf{0} & \textbf{1} & \textbf{2} & \textbf{3} & \textbf{4} & \textbf{5} & \textbf{6} \\
        \hline
        \rowlabel{\textit{A}} & \stagecell{DISP}{Dispatch} & \stagecell{EX}{Fire} & \stagecell{EX}{Miss wait} & \stagecell{EX}{Miss wait} & \stagecell{EX}{Miss wait} & \stagecell{EX}{Miss wait} & \stagecell{CDB}{Complete} \\
        \hline
        \rowlabel{\textit{B}} & \stagecell{DISP}{Dispatch} & \stagecell{RS}{Wait on \textit{A}} & \stagecell{RS}{Wait on \textit{A}} & \stagecell{RS}{Wait on \textit{A}} & \stagecell{RS}{Wait on \textit{A}} & \stagecell{RS}{Wait on \textit{A}} & \stagecell{RS}{Ready after \textit{A}} \\
        \hline
        \rowlabel{\textit{C}} & \stagecell{DISP}{Dispatch} & \stagecell{EX}{Fire} & \stagecell{EX}{Hit wait} & \stagecell{EX}{Hit wait} & \stagecell{CDB}{Complete} &  &  \\
        \hline
        \rowlabel{\textit{D}} & \stagecell{DISP}{Dispatch} & \stagecell{RS}{Wait on \textit{C}} & \stagecell{RS}{Wait on \textit{C}} & \stagecell{RS}{Wait on \textit{C}} & \stagecell{RS}{Wake on \textit{C}} & \stagecell{EX}{Fire} & \stagecell{CDB}{Complete after \textit{A}} \\
        \hline
        \rowlabel{CDB} &  &  &  &  & \stagecell{CDB}{\textit{C}} &  & \stagecell{CDB}{\textit{A} then \textit{D}} \\
        \hline
    \end{tabular}}
    \caption{Compact cycle table for the four-instruction Tomasulo trace. The table emphasizes logical ordering rather than exact micro-timing details.}
    \label{fig:ilp-tomasulo-cycle-table}
\end{figure}
\begin{description}
    \item[Cycle 0.] \textit{A}, \textit{B}, \textit{C}, and \textit{D} dispatch into reservation stations.
    Sources that are already available are copied in as values; missing sources are recorded as tags.
    Destinations are renamed, and the corresponding destination entries are marked not ready.
    \item[Cycle 1.] \textit{A} fires, and independent instruction \textit{C} can also fire because it does not depend on \textit{A}\@.
    This is the key gain from out-of-order scheduling: an older miss does not automatically freeze younger independent work.
    \item[Cycles 2--3.] Nothing new completes.
    \textit{B} is still waiting for \textit{A}, and \textit{D} is still waiting for \textit{C}\@.
    \item[Cycle 4.] \textit{C} completes first and broadcasts its tag and value on the common data bus (CDB).
    Reservation stations compare that tag against all waiting operands, so \textit{D} wakes up and becomes ready to fire.
    \item[Cycle 5.] \textit{D} fires and begins its add using the value from \textit{C}\@.
    \item[Cycle 6.] \textit{A} finally completes its miss and \textit{D} also wants to complete.
    With only one CDB, both cannot broadcast at once, so the usual rule is to let the older instruction win.
    \textit{A} therefore broadcasts first, and \textit{B} immediately becomes ready.
    \textit{D} then broadcasts its own result, and \textit{B} can fire as soon as the machine allows post-broadcast wakeup.
\end{description}
This trace shows why it is important to distinguish \textit{fire} from \textit{complete}.
Firing sends operands into an FU and begins execution.
Completing publishes the result so the rest of the machine can observe it.

\paragraph{Non-blocking caches}\label{par:ilp-tomasulo-nonblocking}
This Tomasulo-style example also motivates a \textit{non-blocking cache}.
Such a cache is often described as supporting \textit{hit under miss}.
If A misses in the cache, the machine does not necessarily want every younger memory access to stop.
Instead, independent cache accesses may continue to probe the cache pipeline.
That allows C to hit and complete while A is still waiting on a lower-level response.
That improves throughput, but it also makes control more complex because the cache and scheduler must now track multiple outstanding requests correctly.


\section{Result Buses and Physical Register Files}\label{sec:ilp-result-bus-prf}
Basic Tomasulo has an obvious bottleneck: if all completed instructions must use one common data bus (CDB), then only one completion can be broadcast in a cycle.
That limitation can be summarized as Eq.~\eqref{eq:ilp-result-bus-bound}.
\begin{equation}\label{eq:ilp-result-bus-bound}
    IPC_{complete} \le N_{result\ buses}.
\end{equation}
Adding more result buses raises the completion ceiling, but it also increases wakeup complexity because each reservation station must compare against more simultaneous broadcasts.

One historical way to manage this cost is to redesign Tomasulo around an explicit physical register file and a RAT rather than storing all transient values only in reservation-station entries.
With physical registers, the scheduler can often replace some associative tag searches with direct lookup by physical register number.
That is faster and usually lower power than asking every waiting instruction to compare against every possible producer all the time.

The tradeoff is a heavier register file.
If many reservation-station entries are checked each cycle, the physical register file may need substantial read bandwidth.
A common first-order estimate is shown in Eq.~\eqref{eq:ilp-prf-port-estimate}.
\begin{equation}\label{eq:ilp-prf-port-estimate}
    N_{read\ ports}\approx 2N_{sched},\qquad
    N_{write\ ports}\approx N_{result\ buses}.
\end{equation}
The factor of two on the read side comes from the common case of two source operands per instruction.
Real designs use banking, clustering, bypass paths, operand caching, and other tricks to reduce this raw multi-port cost, but the estimate is a useful baseline for understanding why renamed out-of-order cores are expensive.


\section{RATs, Ready Bits, and Dispatch in Renamed Machines}\label{sec:ilp-rat-ready}
Once the machine uses an explicit physical register file, dispatch must translate instruction set architecture (ISA) register names into physical register numbers.
That basic renaming step is summarized by Eq.~\eqref{eq:ilp-rat-mapping}.
\begin{equation}\label{eq:ilp-rat-mapping}
    P_{src1}=RAT[A_{src1}],\qquad
    P_{src2}=RAT[A_{src2}],\qquad
    RAT[A_{dest}]\leftarrow P_{free}.
\end{equation}
The first two mappings identify the current physical locations of the source values.
The third mapping allocates a fresh physical destination so younger instructions depend on the new producer rather than on an older architectural name.

In this style of design, the fire rule is usually expressed with \textit{ready bits} on physical registers rather than with \textit{busy bits} on architected registers.
The same logical condition is shown in Eq.~\eqref{eq:ilp-ready-fire}.
\begin{equation}\label{eq:ilp-ready-fire}
    fire_i \Leftarrow ready(P_{src1})\wedge ready(P_{src2})\wedge free(FU_i).
\end{equation}
Each scheduling-queue entry is a reservation station that records the FU type, the destination physical register or tag, and either ready source values or the physical-register identities that will eventually supply them.
When a producing instruction completes, the result bus updates the physical register file and wakes any waiting entries that need that value.

The speculative physical-register state may be called the \textit{future file} or the \textit{messy register file}.
In either case, the term refers to a speculative state that may be ahead of the last coherent architectural state.
The key idea is the same: the speculative state can move ahead aggressively, while the committed state must remain precise.


\section{Precise State and Recovery}\label{sec:ilp-precise-state}
Aggressive out-of-order execution creates a second problem beyond scheduling: the machine may complete instructions in an order that does not match the sequential state the programmer expects.
If an interrupt, page fault, or other exception occurs at the wrong moment, software still needs a coherent restart point.
That requirement is usually stated as the \textit{precise exception} property in Eq.~\eqref{eq:ilp-precise-state}.
\begin{equation}\label{eq:ilp-precise-state}
    \begin{aligned}
        \text{precise exception at }i \Rightarrow\ &\text{all instructions older than }i\text{ are committed}\\
        &\wedge\ \text{no instruction younger than }i\text{ is architecturally visible}
    \end{aligned}
\end{equation}

This requirement was not always imposed in the same way.
In Tomasulo's original floating-point setting on the IBM System/360 Model 91, the out-of-order machinery covered a narrower part of the processor, and the machine could tolerate \textit{imprecise exceptions} in situations that modern general-purpose cores cannot.
Once out-of-order execution extended to ordinary integer, load/store, and control-flow instructions under virtual memory, recovery to an exact sequential state became mandatory rather than optional.


\section{Reorder Buffers and Retirement}\label{sec:ilp-rob}
The standard answer to precise-state recovery is the \textit{reorder buffer (ROB)}.
Dispatch allocates ROB entries in program order.
Execution may finish out of order, but retirement occurs in order from the ROB head, which creates a clean boundary between speculative work and architecturally committed state.

A typical ROB entry stores destination information, a result field, a ready bit, an exception flag, and enough control information to recover the correct program counter (PC)\@.
In simple form, retirement follows Eq.~\eqref{eq:ilp-rob-retire}.
\begin{equation}\label{eq:ilp-rob-retire}
    retire \Leftarrow ready(ROB_{head})\wedge \neg exception(ROB_{head}).
\end{equation}
If the machine can retire at most $W_{retire}$ instructions per cycle, then one cycle may commit up to the oldest $W_{retire}$ consecutive ready non-excepting entries from the ROB head, but it may not skip over an older unready or excepting entry.
Retire width therefore limits architecturally visible IPC even when execution hardware has already completed more work internally.
If the instruction at the ROB head is marked as an exception, the machine instead flushes younger speculative work and transfers control to the retry or handler PC, as shown in Eq.~\eqref{eq:ilp-rob-exception-rule}.
\begin{equation}\label{eq:ilp-rob-exception-rule}
    \begin{aligned}
        exception(ROB_{head}) \Rightarrow\ &\text{flush younger instructions}\\
        &\wedge\ \text{transfer control to the retry or handler PC}
    \end{aligned}
\end{equation}
For example, a page-faulting load is usually retried after the operating system resolves the fault, whereas a trap-like event may resume at the following instruction.

\paragraph{ROB with bypass or forwarding}\label{par:ilp-rob-bypass}
Some designs let dispatch search the ROB from youngest to oldest for the most recent in-flight producer of a source register.
If the matching entry is ready, dispatch can copy the value directly.
If it exists but is not ready, dispatch copies the tag or ROB identity and waits.
If no matching entry exists, the value must already have retired, so dispatch reads it from the architected register file.
This is often described as a ROB with bypass or forwarding.
It is effective, but the search itself is associative and can become part of the critical timing path.

\paragraph{Worked tag-based ROB-bypass trace}\label{par:ilp-rob-bypass-worked}
The following tag-based example makes the lookup rule explicit.
Assume the architected register file initially holds committed ready values for $R1$, $R2$, and $R3$, and that dispatch allocates ROB tags in program order.
Consider:

{\small
\begin{center}
\begin{tabular}{@{}ll@{}}
    \textit{A}: & LW $R1,0(R3)$ \\
    \textit{B}: & ADD $R3,R2,R1$ \\
    \textit{C}: & ADD $R1,R3,R2$
\end{tabular}
\end{center}}

At dispatch, the source resolution is:

{\small
\begin{center}
\renewcommand{\arraystretch}{1.15}
\begin{tabular}{|c|c|c|c|}
    \hline
    Instr.
    & Dest reg/tag & Src \#1 at dispatch & Src \#2 at dispatch \\
    \hline
    \textit{A} & $R1$/A & $R3$ value from architected RF & n/a \\
    \hline
    \textit{B} & $R3$/B & $R2$ value from architected RF & waits on tag A from ROB entry \textit{A} \\
    \hline
    \textit{C} & $R1$/C & waits on tag B from ROB entry \textit{B} & $R2$ value from architected RF \\
    \hline
\end{tabular}
\end{center}}

The key rule is that dispatch always takes the newest in-flight producer.
So \textit{B} does not read the old committed $R1$ value; it waits on tag A\@.
Likewise, \textit{C} does not read the old committed $R3$ value; it waits on tag B\@.
After all three instructions have dispatched, the in-flight structures look like Table~\ref{tab:ilp-rob-bypass-worked}.

\begin{table}[t]
    \centering
    \small
    \renewcommand{\arraystretch}{1.15}
    \begin{tabular}{|c|c|c|c|c|}
        \hline
        Instr.
        & Dest reg/tag & Src \#1 & Src \#2 & Queue status \\
        \hline
        \textit{A} & $R1$/A & $R3$ value ready & n/a & ready to fire \\
        \hline
        \textit{B} & $R3$/B & $R2$ value ready & tag A not ready & waiting on \textit{A} \\
        \hline
        \textit{C} & $R1$/C & tag B not ready & $R2$ value ready & waiting on \textit{B} \\
        \hline
    \end{tabular}

    \vspace{1ex}

    \begin{tabular}{|c|c|c|c|}
        \hline
        ROB entry & Dest reg & Ready & Value source \\
        \hline
        \textit{A} & $R1$ & 0 & pending load result \\
        \hline
        \textit{B} & $R3$ & 0 & pending add result \\
        \hline
        \textit{C} & $R1$ & 0 & pending add result \\
        \hline
    \end{tabular}
    \caption{Tag-based Tomasulo plus ROB bypass after dispatch of \textit{A}, \textit{B}, and \textit{C}.}
    \label{tab:ilp-rob-bypass-worked}
\end{table}

From this point onward, execution follows the dependence chain directly.
\textit{A} completes first and marks ROB entry A ready, which wakes \textit{B}.
\textit{B} then completes and marks ROB entry B ready, which wakes \textit{C}.
\textit{C} completes last.
Retirement still occurs strictly in order \textit{A}, then \textit{B}, then \textit{C}, so after the last retirement both the scheduling queue and the ROB are empty.
After retirement, the architected register file holds the retired results, with $R3$ containing the value produced by \textit{B} and $R1$ containing the younger value produced by \textit{C}.
For any such trace, once the total elapsed cycles are known, fragment IPC is retired instructions divided by elapsed cycles over the chosen measurement interval.

\paragraph{Freeing physical registers}\label{par:ilp-freeing-pregs}
Renamed machines also need a rule for reclaiming old physical registers.
The exact policy differs across implementations, but the underlying principle is simple: an old physical register cannot return to the free list until no older in-flight instruction can still need it.
Common implementations describe this with use counters, pending-free bits, or retirement-based reclamation.
For example, if an architected register was formerly mapped to physical register $P23$ and rename later redirects that architected name to $P27$, the old $P23$ mapping cannot be recycled immediately.
It becomes reclaimable only after no older instruction can still read it and the newer mapping has safely become the committed architectural version.
The details vary, but all correct schemes delay reuse until the older mapping is no longer architecturally or speculatively required.

\paragraph{When architected registers are actually written}\label{par:ilp-areg-write-timing}
Completion and retirement are not the same event.
Completion usually writes the speculative destination into the physical register file and marks the matching ROB entry ready.
The architected register file is updated only when that ROB entry reaches the head and retires in order.
This separation is what allows the machine to execute aggressively while still presenting precise sequential state to software.

\paragraph{Worked PRegs/RAT/ROB end-state trace}\label{par:ilp-preg-rob-worked}
The following short trace makes the bookkeeping easier to follow.
Assume the initial mappings are $R0\rightarrow P0$, $R1\rightarrow P1$, $R2\rightarrow P2$, and $R3\rightarrow P3$, with $P4$ through $P7$ free.
Assume two-wide fetch and dispatch, two-wide retirement, two 1-cycle ALUs, one 1-cycle LSU, and a back-to-front cycle evaluation so that Execute can mark a destination physical register ready before Schedule runs later in the same cycle.
Consider the following sequence:

{\small
\begin{center}
\begin{tabular}{@{}ll@{}}
    \textit{A}: & LW $R3,0(R1)$ \\
    \textit{B}: & ADD $R1,R3,R2$ \\
    \textit{C}: & ADD $R2,R2,R1$
\end{tabular}
\end{center}}

One legal cycle-by-cycle interpretation is:
\begin{description}
    \item[Cycle 1.] Fetch \textit{A} and \textit{B}.
    \item[Cycle 2.] Dispatch \textit{A} and \textit{B}. \textit{A} allocates $P4$ as the new destination for $R3$ and records old mapping $P3$ in its ROB entry. \textit{B} therefore reads $P4$ as its first source, allocates $P5$ as the new destination for $R1$, and records old mapping $P1$ in its ROB entry.
    Fetch \textit{C}.
    \item[Cycle 3.] \textit{A} fires on the LSU. Dispatch \textit{C}, which now reads $P2$ and $P5$, allocates $P6$ as the new destination for $R2$, and records old mapping $P2$ in its ROB entry.
    \item[Cycle 4.] \textit{A} completes, marking $P4$ ready and its ROB entry ready.
    Because that completion is visible before scheduling in the same cycle, \textit{B} can now fire.
    \item[Cycle 5.] \textit{A} retires, so the architected copy of $R3$ now comes from $P4$ and old mapping $P3$ can be freed. \textit{B} completes, marking $P5$ ready, and \textit{C} can then fire.
    \item[Cycle 6.] \textit{B} retires, so the architected copy of $R1$ now comes from $P5$ and old mapping $P1$ can be freed. \textit{C} completes, marking $P6$ ready.
    \item[Cycle 7.] \textit{C} retires, so the architected copy of $R2$ now comes from $P6$ and old mapping $P2$ can be freed.
\end{description}

The exact cycle numbers can shift by one under a different but internally consistent simulator ordering, but the final renamed state does not change.

{\small
\begin{center}
\renewcommand{\arraystretch}{1.15}
\begin{tabular}{|c|c|c|c|c|c|}
    \hline
    Instr.
    & FU & Dest physical reg. & Src \#1 & Src \#2 & Final status \\
    \hline
    \textit{A} & LSU & $P4$ & $P1$ & n/a & completed and retired \\
    \hline
    \textit{B} & ALU & $P5$ & $P4$ & $P2$ & completed and retired \\
    \hline
    \textit{C} & ALU & $P6$ & $P2$ & $P5$ & completed and retired \\
    \hline
\end{tabular}

\vspace{1ex}

\begin{tabular}{|c|c|c|c|}
    \hline
    ROB entry & Dest architected reg. & Prev physical reg. & Final status \\
    \hline
    \textit{A} & $R3$ & $P3$ & retired \\
    \hline
    \textit{B} & $R1$ & $P1$ & retired \\
    \hline
    \textit{C} & $R2$ & $P2$ & retired \\
    \hline
\end{tabular}

\vspace{1ex}

\begin{tabular}{|c|c|}
    \hline
    Architected reg. & Final physical reg. \\
    \hline
    $R0$ & $P0$ \\
    \hline
    $R1$ & $P5$ \\
    \hline
    $R2$ & $P6$ \\
    \hline
    $R3$ & $P4$ \\
    \hline
\end{tabular}
\hspace{1.5em}
\begin{tabular}{|c|c|c|}
    \hline
    Physical reg. & Ready & Free \\
    \hline
    $P0$ & 1 & 0 \\
    \hline
    $P1$ & 1 & 1 \\
    \hline
    $P2$ & 1 & 1 \\
    \hline
    $P3$ & 1 & 1 \\
    \hline
    $P4$ & 1 & 0 \\
    \hline
    $P5$ & 1 & 0 \\
    \hline
    $P6$ & 1 & 0 \\
    \hline
    $P7$ & 0 & 1 \\
    \hline
\end{tabular}
\end{center}}

Three rules determine the outcome.
Dispatch allocates the next free physical destination and records the previous architectural mapping in the ROB\@.
Completion makes the speculative destination ready for younger consumers.
Retirement is the event that both updates the architected view and frees the previous physical mapping.


\section{Future Files and Checkpoint/Repair}\label{sec:ilp-future-file-checkpoint}
One useful way to describe renamed out-of-order state is to separate it into two layers.
The \textit{future file} holds the most recent speculative values that execution is building toward.
The \textit{architected register file} holds only committed state.
Execution and wakeup consult the speculative state, while retirement copies correct in-order results into the architected state.
This explains how the machine can behave aggressively internally while still presenting a sequential interface to software.
In a physical-register implementation, this often means retirement copies the committed value from the current physical register into the architected register file, while exception recovery can restore a precise view by redirecting the register alias table (RAT) back to the committed architectural mappings.

\paragraph{Comparison with tag-only Tomasulo}\label{par:ilp-tag-vs-preg}
A physical-register plus RAT design does not automatically increase the amount of ILP that the scheduler can discover.
If a tag-based Tomasulo machine and a physical-register machine have the same function units, the same wakeup/select policy, the same retire width, and enough result buses that completion bandwidth is not the bottleneck, then they can expose the same execution schedule and the same IPC on a given trace.
The main differences are in bookkeeping cost, lookup style, recovery organization, and physical-register reuse rather than in the dependence constraints themselves.
Performance diverges only when one design changes a practical limit such as result-bus count, port bandwidth, wakeup timing, or retirement bandwidth.

The ROB is not the only recovery method.
\textit{Checkpoint/repair} schemes periodically capture a coherent snapshot of register state and use that snapshot as the recovery point.
One common description uses three conceptual views of the state: a messy current copy, a working copy being built toward coherence, and a golden copy representing the last known coherent sequential state.

The key control concept is the \textit{instruction boundary}.
Dispatch chooses a dynamic boundary, often denoted by markers such as IB1 and IB2.
Instructions that complete before the current boundary update both the messy state and the working copy.
Instructions that complete after the boundary update only the messy state.
Once every instruction older than the boundary has completed, the working copy has become coherent and may be promoted to the new golden copy.
The machine then chooses a new boundary and repeats the process.

If an exception occurs, the machine restores the golden copy and continues from that last coherent point.
It may then re-execute younger instructions sequentially or under tighter control until it reaches the faulting instruction and handles the event precisely.
This can reduce some of the timing pressure associated with aggressive ROB-bypass logic, though it introduces its own bookkeeping and copying costs.


\section{Speculation Without and With Prediction}\label{sec:ilp-speculation}
Superscalar pipelines use speculation to keep FUs busy.
Some designs can speculate without high-quality prediction (for example, eager execution ideas), but the useful work fraction is then limited.
High-accuracy branch handling usually gives better effective IPC because fewer speculative instructions are squashed.


\section{Security Implications of Speculative Execution}\label{sec:ilp-security}
Precise exceptions solve the architectural recovery problem, but they do not automatically solve the microarchitectural side-effect problem.
An instruction that is later squashed may still have changed caches, predictors, or other timing-visible structures before the machine discovered that the instruction should not retire.
That is the core issue behind speculative side channels.

The best-known examples are Spectre- and Meltdown-style attacks.
At a high level, the attacker causes the processor to execute work that will later be squashed or faulted.
Even though the architectural state rolls back correctly, the speculative work may still have fetched a cache line or trained a predictor.
The attacker then measures timing differences to infer information that should never have become visible through the ISA\@.

Shared-core execution can make this worse.
With \textit{hyperthreading}, or more generally simultaneous multithreading, two threads share parts of one physical core.
That increases the opportunity for one thread to observe side effects caused by another thread's speculative activity.
The security lesson is broader than any single named vulnerability: microarchitectural cleanup matters, not just architectural rollback.


\section{Implementation Pressure}\label{sec:ilp-implementation-pressure}
Once out-of-order execution moved from isolated examples into full commercial processors, these mechanisms had to satisfy real timing, area, power, and verification constraints.
That practical pressure is one reason the field explored multiple recovery styles rather than converging immediately on a single universal design.

The architectural lesson is that elegant scheduling ideas are not sufficient by themselves.
ROB bypass logic, associative searches, speculative-state tracking, and recovery control all stress timing closure and verification.
As out-of-order windows grow wider and deeper, state management and recovery logic become first-order design problems rather than secondary implementation details.


\section{Interpreting Out-of-Order Processor Statistics}\label{sec:ilp-sim-stats}
Trace-driven superscalar simulators often report more than cycles and IPC\@.
Those counters are useful only when their meanings and denominators are clear.

\paragraph{Fetched versus retired instructions}\label{par:ilp-fetched-vs-retired}
The architecturally meaningful IPC uses retired instructions, not fetched instructions.
\textit{Trace instructions} count the dynamic instructions in the workload itself.
\textit{Fetched instructions} count every instruction the front end brought in, including wrong-path work fetched before a branch misprediction was repaired.
\textit{Retired instructions} count only the instructions that reached precise architectural state.
In a correct full run, retired instructions usually match the trace instruction count, while fetched instructions may be much larger when prediction accuracy is poor.

\paragraph{Miss and misprediction counters}\label{par:ilp-miss-mispredict-counters}
Branch-misprediction counts measure control-flow repair events.
Instruction-cache misses reduce front-end supply, while data-cache misses usually delay load or store completion and can starve dependent instructions.
These counters therefore explain why IPC may fall even when the machine nominally has ample fetch width and many function units.

\paragraph{Queue-pressure counters}\label{par:ilp-queue-pressure}
Maximum and average dispatch-queue, scheduling-queue, and ROB occupancy show where pressure accumulates.
If ROB usage remains near its limit and dispatch is frequently blocked because the ROB is full, then commit bandwidth, long-latency instructions, or recovery delays are throttling forward progress.
If scheduling queues remain lightly used but \textit{no-fire cycles} are high, then the machine is more likely waiting on ready operands or memory completion than on queue capacity.

\paragraph{Isolating control and memory effects}\label{par:ilp-isolate-effects}
Many simulators also offer a no-miss or no-misprediction mode.
Such a configuration does not model real execution, but it is analytically useful because it isolates pure width, scheduling, and dependence limits from penalties caused by cache misses and poor control-flow prediction.

\section{Chapter 6 Summary}\label{sec:chapter6-summary-ilp}
Chapter 6 develops dynamic instruction-level parallelism as a concrete microarchitectural path to $CPI<1$.
Superscalar design achieves this by widening the pipeline's effective throughput and managing dependencies, hazards, and speculative state carefully.
FICO-style scoreboarding illustrates the basic mechanism, while renaming, Tomasulo-style wakeup and result broadcast, and stronger out-of-order policies push performance further.
Once the machine executes and completes work out of order, however, scheduling is no longer enough by itself.
The core must also maintain precise architectural state through mechanisms such as physical-register renaming, reorder buffers, future files, or checkpoint/repair schemes.
Modern processors therefore combine ILP techniques with careful recovery, retirement, and microarchitectural side-effect control.
Simulator-visible counters such as fetched-versus-retired instructions, misprediction counts, miss counts, queue occupancy, and ROB pressure then provide a practical way to diagnose which limits are dominating a given design point.
These ideas together form the bridge from classic five-stage thinking to modern high-throughput scalar cores.
